Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-08-24 paper/paper.tex
Finish semantics section

\subsection{Semantics}
Formulae in \textbf{MLSS} contain arbitrarily nested set terms.
We account for this arbitrary nesting by basing the semantics on the von Neumann hierarchy $\mathcal{V}$ of sets that are defined inductively as
  \[
    \begin{array}{rclr}
      \mathcal{V}_0 & = & \emptyset, \\
      \mathcal{V}_{\alpha + 1} & = & \mathcal{P}(\mathcal{V}_\alpha) & \text{for each ordinal $\alpha$}, \\
      \mathcal{V}_{\lambda} & = & \bigcup_{\mu < \lambda} \mathcal{V}_\mu & \text{for each limit ordinal $\lambda$}, \\
      \mathcal{V} & = & \bigcup_\alpha \mathcal{V}_\alpha, \\
    \end{array}
  \]
where $\mathcal{P}(S)$ is the powerset of $S$.
The sets in $\mathcal{V}$ are well-founded meaning that there can be no membership cycle in $\mathcal{V}$.

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitiation is provided by Lawrence in an AFP entry~\cite{zfc_in_hol_afp}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
An article by Lawrence~\cite{zfc_in_hol_afp} provides some further context to this formalisation.

\begin{lstlisting}
fun I$_\text{st}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  I$_\text{st}$ M $\emptyset$
| I$_\text{st}$ M (Var x) = M x
| I$_\text{st}$ M (Single s) = vset {I$_\text{st}$ M s}
| I$_\text{st}$ M (s $\sqcup_\text{s}$ t) = I$_\text{st}$ M s $\sqcup$ I$_\text{st}$ M t 
| I$_\text{st}$ M (s $\sqcap_\text{s}$ t) = I$_\text{st}$ M s $\sqcap$ I$_\text{st}$ M t 
| I$_\text{st}$ M (s $-_\text{s}$ t) = I$_\text{st}$ M s $-$ I$_\text{st}$ M t 
\end{lstlisting}

\begin{lstlisting}
fun I$_\text{sa}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ V where
  I$_\text{sa}$ M (s $\in_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ M s $\in$ elts (I$_\text{st}$ M t)
| I$_\text{sa}$ M (s $\approx_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ M s = I$_\text{st}$ M t
\end{lstlisting}

>>>>

\subsection{Semantics}
Formulae in \textbf{MLSS} contain arbitrarily nested set terms.
We account for this arbitrary nesting by basing the semantics on the von Neumann hierarchy $\mathcal{V}$ of sets that is defined inductively as
  \[
    \begin{array}{rclr}
      \mathcal{V}_0 & = & \emptyset, \\
      \mathcal{V}_{\alpha + 1} & = & \mathcal{P}(\mathcal{V}_\alpha) & \text{for each ordinal $\alpha$}, \\
      \mathcal{V}_{\lambda} & = & \bigcup_{\mu < \lambda} \mathcal{V}_\mu & \text{for each limit ordinal $\lambda$}, \\
      \mathcal{V} & = & \bigcup_\alpha \mathcal{V}_\alpha, \\
    \end{array}
  \]
where $\mathcal{P}(S)$ is the powerset of $S$.
The sets in $\mathcal{V}$ are well-founded meaning that there can be no membership cycle in $\mathcal{V}$.

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitiation is provided by Lawrence in an AFP entry~\cite{zfc_in_hol_afp}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
An article by Lawrence~\cite{zfc_in_hol} provides some further context to this entry.
The entry declares a type \lstinline!V! that is equipped with a function \lstinline!elts :: V $\Rightarrow$ V set! that maps a set to its elements.
Even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class.
Only if the class is small enough, as indicated by the polymorphic predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, it is indeed a set.
Inversely, this predicate characterises those sets that can be turned into a set of type \lstinline!V! using the polymorphic function \lstinline!vset :: 'a set $\Rightarrow$ V!.
The type \lstinline!V! comes with all the usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$).
The empty set coincides with the ordinal $0$ so we denote it with \lstinline!0 :: V!.

Equipped with the above, we define the interpretation function \lstinline!I$_\text{st}$! for set terms that interprets a set term with respect to a valuation function \lstinline!M :: 'a $\Rightarrow$ V! for variable.
\begin{lstlisting}
fun I$_\text{st}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  I$_\text{st}$ M $\emptyset$ = 0
| I$_\text{st}$ M (Var x) = M x
| I$_\text{st}$ M (Single s) = vset {I$_\text{st}$ M s}
| I$_\text{st}$ M (s $\sqcup_\text{s}$ t) = I$_\text{st}$ M s $\sqcup$ I$_\text{st}$ M t 
| I$_\text{st}$ M (s $\sqcap_\text{s}$ t) = I$_\text{st}$ M s $\sqcap$ I$_\text{st}$ M t 
| I$_\text{st}$ M (s $-_\text{s}$ t) = I$_\text{st}$ M s $-$ I$_\text{st}$ M t 
\end{lstlisting}
The interpretation function \lstinline!I$_\text{sa}$! for set atoms is straightforward as well.
\begin{lstlisting}
fun I$_\text{sa}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ V where
  I$_\text{sa}$ M (s $\in_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ M s $\in$ elts (I$_\text{st}$ M t)
| I$_\text{sa}$ M (s $\approx_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ M s = I$_\text{st}$ M t
\end{lstlisting}
For formulae, we write \lstinline!M $\models$ $\phi$! for the judgement that $\phi$ holds under the valuation function \lstinline!M!.
We omit the implementation of $\models$ for brevity.
